---
title: "Lab 07 — t-tests and ANOVA"
author: "EAES 480 — Dylan Enwia Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

> GM: 9.75/10 - Excellent — Thoughtful, well-organized, and hits all key points. 

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends hypothesis testing from **claims about one mean** to:

- **one-sample t-tests**
- **two-sample t-tests**
- **paired t-tests**
- **ANOVA**

You will use the **US-AMS simplified AmeriFlux dataset** again and compare means across different time-based groups.

The core logic stays the same:

> estimate → uncertainty → test statistic → p-value

This lab emphasizes both **manual understanding** and **automated testing in R**.

---

# Learning goals

By the end of this lab, you should be able to:

- Compute and interpret a **one-sample t-test**
- Distinguish between **two-sample** and **paired** t-tests
- Compute and interpret **degrees of freedom**
- Explain the structure of an **ANOVA**
- Compare manual calculations to `t.test()` and `aov()`
- Use visualizations to support statistical interpretation
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=TRUE}
# GOAL: Read the dataset and keep 2022–2023.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local %in% c(2022, 2023))

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=TRUE}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

---

# Part I — One-sample t-test

## 1) Choose a day and a variable

Select:
- one **year**
- one **numeric variable**

Examples:
- `ta`
- `swc`
- `le`
- `fc`

```{r choose_one_sample, echo=TRUE, eval=TRUE}
target_year <- 2023
target_var  <- "ta"

sample_data <- df %>%
  filter(year_local == target_year) %>% 
  slice_sample(n = 50)

x <- sample_data[[target_var]]

mean(x, na.rm = TRUE); sd(x, na.rm = TRUE); length(x)
```

> What variable and day did you choose, and why?

I used air temperature (ta) in 2023 and used a sample of 50 observations. I chose ta because it is easy to interpret across time periods especially during different seasons.

## 2) State a null hypothesis

Write a claim about the mean of your chosen variable on that day.

Examples:
- “The mean temperature on this day is 15 °C.”
- “The mean latent heat flux on this day is 100 W m$^{-2}$.”

Fill in:

> $H_0$: `The mean air temperature for this sampled day is 13 °C`

> $H_A$: `The mean air temperature for this sampled day is not 13 °C`

---

## 3) Compute the one-sample t-statistic manually

Use:

$t = (\bar{x} - \mu_0) / (s / \sqrt{n})$

```{r one_sample_manual, echo=TRUE, eval=TRUE}
mu0 <- 13

xbar <- mean(x, na.rm = TRUE)
s <- sd(x, na.rm = TRUE)
n <- sum(!is.na(x))

t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

---

## 4) Compute the p-value and degrees of freedom

```{r one_sample_pvalue, echo=TRUE, eval=TRUE}
df_t <- n - 1

p_value <- 2 * pt(-abs(t_stat), df = df_t)

tibble(
  xbar = xbar,
  s = s,
  n = n,
  df_t = df_t,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the p-value suggest about your null hypothesis?
The p-value is 0.835, which is larger than the typical significance level of 0.05. This means it fails to reject the null hypothesis.

> GM: Good answer — Clear fail-to-reject conclusion.

---

## 5) Compare with `t.test()`

```{r one_sample_ttest, echo=TRUE, eval=TRUE}
t.test(x, mu = mu0)
```

> Does the `t.test()` output agree with your manual calculation? If not, why might that be?
 Yes, the t.test() output agrees with the manual calculation.
---

# Part II — Two-sample t-test

## 6) Compare the same variable across two day ranges

Choose:
- the **same variable**
- two different **day ranges in 2023**

```{r choose_two_sample, echo=TRUE, eval=TRUE}
x1 <- df %>%
  filter(year_local == 2023, doy > 20 & doy < 22) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 180 & doy < 182) %>%
  pull(target_var)

mean(x1, na.rm = TRUE); sd(x1, na.rm = TRUE); length(x1)
mean(x2, na.rm = TRUE); sd(x2, na.rm = TRUE); length(x2)
```

> Why did you choose these two days?
I chose these two day ranges because they represent very different parts of the year. One in the summer and one in winter. For air temperature, this should create a contrast in means between these two set of days.
---

## 7) Summarize the difference in means

```{r two_sample_summary, echo=TRUE, eval=TRUE}
xbar1 <- mean(x1, na.rm = TRUE)
xbar2 <- mean(x2, na.rm = TRUE)

s1 <- sd(x1, na.rm = TRUE)
s2 <- sd(x2, na.rm = TRUE)

n1 <- sum(!is.na(x1))
n2 <- sum(!is.na(x2))

mean_diff <- xbar1 - xbar2

tibble(
  xbar1 = xbar1,
  xbar2 = xbar2,
  mean_diff = mean_diff
)
```

---

## 8) Compute the two-sample t-statistic manually (Welch)

```{r two_sample_manual, echo=TRUE, eval=TRUE}
se_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))

t_stat <- mean_diff / se_diff

df_welch <- ((s1^2 / n1 + s2^2 / n2)^2) /
  (((s1^2 / n1)^2 / (n1 - 1)) + ((s2^2 / n2)^2 / (n2 - 1)))

p_value <- 2 * pt(-abs(t_stat), df = df_welch)

tibble(
  mean_diff = mean_diff,
  se_diff = se_diff,
  t_stat = t_stat,
  df_welch = df_welch,
  p_value = p_value
)
```

> What does the sign of `mean_diff` mean? What does the p-value mean?
The negative mean_diff shows that the first group has a lower mean than the second group. The p-value of 5.19e-107 is smaller than 0.05, so thus we can reject the null hypothesis. Overall we can interpret that the two groups are different from these two values. 
---

## 9) Compare with `t.test()`

```{r two_sample_ttest, echo=TRUE, eval=TRUE}
t.test(x1, x2)
```

> Does the automated t-test agree with your manual result?
Yes, the automated t-test agrees with the manual result.
---

## 10) Visualize the two groups

```{r two_sample_plot, echo=TRUE, eval=TRUE}
day_range1 <- "20_22"
day_range2 <- "180_182"

plot_df <- bind_rows(
  tibble(group = paste0("DOY_", day_range1), value = x1),
  tibble(group = paste0("DOY_", day_range2), value = x2)
)

ggplot(plot_df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = target_var,
    title = "Two-sample comparison"
  )
```

> Does the plot match your statistical conclusion?
Yes the plot does match my statistical conclusion. We can see the two groups clearly separated with basically no overlap, which shows a large difference in mean_diff. 
---

# Part III — Paired t-test

## 11) Compare the same days across two years

Choose one DOY and compare **2022 vs 2023** for the same variable.

```{r paired_setup, echo=TRUE, eval=TRUE}
x1 <- df %>%
  filter(year_local == 2022, doy > 150 & doy < 156) %>%
  arrange(date) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy >  150 & doy < 156) %>%
  arrange(date) %>%
  pull(target_var)

length(x1); length(x2)
```

> Why is this a paired comparison rather than an independent two-sample comparison?
This is a paired comparison because each observation in 2022 is matched with a corresponding observation in 2023 for the same day of year and ordered by time. Since both datasets have equal length each pair represents the same temporal position across years.
--- 

## 12) Compute paired differences manually

```{r paired_manual, echo=TRUE, eval=TRUE}
d <- x1 - x2

dbar <- mean(d, na.rm = TRUE)
sd_d <- sd(d, na.rm = TRUE)
n <- sum(!is.na(d))

t_stat <- dbar / (sd_d / sqrt(n))
p_value <- 2 * pt(-abs(t_stat), df = n - 1)

tibble(
  dbar = dbar,
  sd_d = sd_d,
  n = n,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the sign of `dbar` mean?
Since dbar is positive it shows the difference between paired observations is greater than zero. This means that the first group the year 2022 has higher values than the second group 2023.
---

## 13) Compare with `t.test(..., paired = TRUE)`

```{r paired_ttest, echo=TRUE, eval=TRUE}
t.test(x1, x2, paired = TRUE)
```

> How does the paired t-test differ conceptually from the two-sample t-test?
A paired t-test compares the mean between matched observations between groups while a two-sample t-test compares the means of two independent groups and doesn't account for the same pairing as seen in the paired t-test.
---

# Part IV — ANOVA

## 14) Create a categorical grouping variable

Group `ta` into broad seasonal categories:

- Summer
- Winter
- Other

```{r make_groups_anova, echo=TRUE, eval=TRUE}
df_anova <- df %>%
  filter(!is.na(ta)) %>%
  mutate(
    month_group = case_when(
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      TRUE ~ "Other"
    )
  )

df_anova %>% count(month_group)
```

---

## 15) Visualize the groups

```{r plot_anova_groups, echo=TRUE, eval=TRUE}
ggplot(df_anova, aes(x = month_group, y = ta, fill = month_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = "ta",
    title = "Air temperature by seasonal group"
  )
```

> Do the groups look different enough that ANOVA might detect a difference?
I would say yes the groups look different enough that ANOVA might detect a difference. I say this because we can see some seperation between the groups and thus ANOVA will likely detect a difference.
---

## 16) Compute ANOVA components manually

```{r anova_manual_components, echo=TRUE, eval=TRUE}
grand_mean <- mean(df_anova$ta, na.rm = TRUE)

group_stats <- df_anova %>%
  group_by(month_group) %>%
  summarise(
    group_mean = mean(ta, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ss_between <- sum(group_stats$n * (group_stats$group_mean - grand_mean)^2)

df_anova2 <- df_anova %>%
  left_join(group_stats, by = "month_group")

ss_within <- sum((df_anova2$ta - df_anova2$group_mean)^2)

k <- n_distinct(df_anova$month_group)
n_total <- nrow(df_anova)

df_between <- k - 1
df_within <- n_total - k

ms_between <- ss_between / df_between
ms_within <- ss_within / df_within

f_stat <- ms_between / ms_within
p_value <- pf(f_stat, df1 = df_between, df2 = df_within, lower.tail = FALSE)

tibble(
  grand_mean = grand_mean,
  ss_between = ss_between,
  ss_within = ss_within,
  df_between = df_between,
  df_within = df_within,
  f_stat = f_stat,
  p_value = p_value
)
```

> In your own words, what do `ss_between` and `ss_within` represent?
Ss_between is the variation in a groups means when compared to the overall mean. From that we can see the differences between groups in the data set. S_swithin is the variation of individual obervsations within different groups. 
---

## 17) Compare with `aov()`

```{r run_anova, echo=TRUE, eval=TRUE}
model <- aov(ta ~ month_group, data = df_anova)
summary(model)
```

> What does the ANOVA p-value tell you? Does it tell you which groups differ?
The ANOVA p-valye tells me that the p-value is small and thus indicates that there is a statistically significant difference among the group means, so we reject the null hypothesis that every groip here is equal. ANOVA doesn't specify which groups differ from each other.
---

## 18) Final reflection

Answer in **5–7 sentences**:

- How is a one-sample t-test different from a two-sample t-test?
- Why is a paired t-test special?
- How is ANOVA a generalization of the t-test?
- What stayed the same across all these tests?

> *A one sample t-test compares a sample mean from one group mean, while a two-sample t-test compares the means of two different independent groups. A paired t-test is special because it can show the differences within matched pairs from the same number of observations. ANOVA is a generalization of the t-test because it can compare more than two groups at a time. Across all tests comparing the differences in means stayed the same. We saw in each test it uses a p-value to determine the statistical significance of a group.*

> GM: -0.25 - Good overall, but a few spots could be more precise (e.g., “one group mean” should be “a hypothesized value”; “comparing the differences in means stayed the same” is a bit vague). Still shows solid understanding.

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
